---
title: "TriangulaR"
output: html_notebook
---

# Setup

```{r}

library(tidyverse)
library(SNPfiltR)
library(vcfR)
library(triangulaR)
library(ggrepel)

```

# Data preparation and exploration

First, reading in the data, which should be in VCF format, HW-filtered and not include any unrelated individuals.

```{r}

data <- read.vcfR("../data/02_3_hwe_filtered_unrelated.vcf", verbose = T)

```

We can also have a quick look:

```{r}

data

```
Another way of exploring our data is to check the SNP library with SNPfiltR:

```{r}

hard_filter(data)

```
Now, setting up a population map and doing some data wrangling:

```{r}

popmap <- read.table("../data/museum_contemporary_samples.tsv", 
                     sep="\t", header=T) %>% 
  mutate(population_full = case_when(
    population == "zlat" ~ "Z. lateralis",
    population == "zlat_hist" ~ "Z. lateralis (historical)",
    population == "zten" ~ "Z. tenuirostris",
    population == "zlat_zten" ~ "Z. lateralis x Z. tenuirostris",
    population == "zalb" ~ "Z. albogularis",
    population == "zlat_zalb" ~ "Z. lateralis x Z. albogularis",
    population == "zten_hist" ~ "Z. tenuirostris (historical)"
  ))

popmap <- popmap %>% select(sample, population_full) %>% 
  rename(id = sample, pop = population_full)

# Remove related individual
popmap <- popmap %>% filter(id!="AEB01_L0004")

# Rename samples in popmap to match vcf
popmap$id <- colnames(data@gt)[-1]

```

# Plotting Z. lateralis x Z. tenuirostris

We first need to remove individuals with Z. albogularis ancestry.

IMPORTANT NOTE: AEB01_L0004 has already been dropped in data prep stage because it was shown to be related to AEB01_L0003. 

```{r}

# Set names of individuals to drop
inds_to_drop <- c("AEB01_L0001", "AEB01_L0007", "AEB01_L0009")

# Get set difference
ind_names <- colnames(data@gt)[-1]
keep_inds <- setdiff(ind_names, inds_to_drop)

# Subset dataset
data_filt_zten <- data[, c(TRUE, colnames(data@gt)[-1] %in% keep_inds)]

# Update the popmap
popmap_filt_zten <- popmap %>% filter(!id %in% inds_to_drop)


```

Now preparing the Z. lateralis x Z. tenuirostris data for plotting:

```{r}

data_filt_zten_diff <- alleleFreqDiff(vcfR = data_filt_zten, pm = popmap_filt_zten, p1 = "Z. lateralis", p2 = "Z. tenuirostris", difference = 0.9)
hi.het <- hybridIndex(vcfR = data_filt_zten_diff, pm = popmap_filt_zten, p1 = "Z. lateralis", p2 = "Z. tenuirostris")

```

Making triangle plot:

```{r}
colours <- c("#E9C46A", "#E9C46A", "#F4A259", "#D6745C", "#D6745C")
zten_plot <- triangle.plot(hi.het, colors=colours) + 
  theme(legend.title=element_blank(),
        legend.position='none')+
  geom_label_repel(aes(label = id), box.padding=0.5, max.overlaps=3,
                   show.legend = F)
zten_plot

ggsave(zten_plot, filename="../plots/triangular/zten_triangle_plink_AEB01_L0004_removed.png", dpi=300, width = 3, height = 3)

```

# Plotting Z. lateralis x Z. albogularis

First, dropping any individuals with Z. tenuirostris ancestry:

```{r}
# Set names of individuals to drop (any with tenuirostris ancestry)
inds_to_drop <- c("AEB01_L0002", "AEB01_L0003", "AEB01_L0005",
                  "AEB01_L0008", "AEB01_L0010",
                  "NOR124_NOR124", "NOR125_NOR125", "NOR132_NOR132", 
                  "NOR134_NOR134", "NOR140_NOR140", "NOR141_NOR141", 
                  "NOR142_NOR142")

# Get set difference
ind_names <- colnames(data@gt)[-1]
keep_inds <- setdiff(ind_names, inds_to_drop)

# Subset dataset
data_filt_zalb <- data[, c(TRUE, colnames(data@gt)[-1] %in% keep_inds)]

# Update the popmap
popmap_filt_zalb <- popmap %>% filter(!id %in% inds_to_drop)

```


Now making the triangle plot (same as above):

```{r}

data_filt_zalb_diff <- alleleFreqDiff(vcfR = data_filt_zalb, pm = popmap_filt_zalb, p1 = "Z. lateralis", p2 = "Z. albogularis", difference = 0.9)
hi.het_zalb <- hybridIndex(vcfR = data_filt_zalb_diff, pm = popmap_filt_zalb, p1 = "Z. lateralis", p2 = "Z. albogularis")
colours <- c("#64ACA3", "#E9C46A",  "#E9C46A", "#A2C75E")

# plotting
zalb_plot <- triangle.plot(hi.het_zalb, colors=colours) +
  theme(legend.title=element_blank(),
        legend.position='none') +
  geom_label_repel(aes(label = id), box.padding=0.5, max.overlaps=1,
                  show.legend = F)
zalb_plot
ggsave(zalb_plot, filename="../plots/triangular/zalb_triangle_plink.png", dpi=300, width = 3, height = 3)


```

Finally, create combined plot with patchwork package:

```{r}

combined <- zten_plot / zalb_plot

ggsave(combined, filename="../plots/combined_triangle_plink.png", dpi=300, width = 5, height = 6)


```

# Check distribution of sites for each species

```{r}

# ------------- z. tenuirostris ----------------

# Extract site characteristics
fix_zten <- as.data.frame(getFIX(data_filt_zten_diff), stringsAsFactors=F)

# Add position and chrom number
fix_zten$POS <- as.numeric(fix_zten$POS)
fix_zten$CHROM <- factor(fix_zten$CHROM, levels = unique(fix_zten$CHROM))

# Calculate chromosome lengths from the data
chr_lengths <- fix_zten %>%
  group_by(CHROM) %>%
  summarise(chr_length = max(POS)) %>%
  arrange(CHROM)

# Calculate cumulative offsets
chr_lengths$offset <- c(0, cumsum(head(chr_lengths$chr_length, -1)))

# Join offsets back to markers
fix_zten <- left_join(fix_zten, chr_lengths, by = "CHROM")

# Genome-wide coordinate
fix_zten$genome_pos <- fix_zten$POS + fix_zten$offset

# Count markers in 1 Mb windows
bins_zten <- fix_zten %>%
  mutate(bin = floor(genome_pos / 1e6) * 1e6) %>%
  count(bin)

distrib_plot_zten <- ggplot(bins_zten, aes(bin / 1e6, n)) +
  geom_col(width = 1,fill="#D6745C") +
  labs(x = "Genome position (Mb)",
       y = "No. AIMs") +
  theme_classic()+
  ggtitle("Z. tenuirostris")

# # ------------- z. albogularis  ----------------

# Extract site characteristics
fix_zalb <- as.data.frame(getFIX(data_filt_zalb_diff), stringsAsFactors=F)

# Add position and chrom number
fix_zalb$POS <- as.numeric(fix_zalb$POS)
fix_zalb$CHROM <- factor(fix_zalb$CHROM, levels = unique(fix$CHROM))

# Join offsets back to markers
fix_zalb <- left_join(fix_zalb, chr_lengths, by = "CHROM")

# Genome-wide coordinate
fix_zalb$genome_pos <- fix_zalb$POS + fix_zalb$offset

# Count markers in 1 Mb windows
bins_zalb <- fix_zalb %>%
  mutate(bin = floor(genome_pos / 1e6) * 1e6) %>%
  count(bin)

distrib_plot_zalb <- ggplot(bins_zalb, aes(bin / 1e6, n)) +
  geom_col(width = 1,fill="#64ACA3") +
  labs(x = "Genome position (Mb)",
       y = "No. AIMs") +
  theme_classic()+
  ggtitle("Z. albogularis")

# Make combined plot

distrib_plot <-
  distrib_plot_zten /
  distrib_plot_zalb

distrib_plot

ggsave(distrib_plot, filename="../plots/triangular_distribution_1mb_windows.png", dpi=300, width = 8, height = 6)


```


